Finite-state morphological transducers for three Kypchak languages

### 2014-02-16 kipchak-lrec-2013/kipchak-paper.tex
update
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@50292 72bbbca6-d526-0410-a7d9-f06f51895060

@@

\begin{map*}[htbp]
	\includegraphics[width=\textwidth]{map/map}
	\caption{The three subbranches of Kypchak (N, S, W), roughly divided with black lines, showing the geographic distribution of the three languages for which transducers were developed (tat, kaz, kum).  Language codes are from ISO 639-3.}
	\label{map1}
\end{map*}

@@

The Kumyk transducer was developed starting at the beginning of October, 2013 as experiment to see how difficult it would be to extend lessons learned from the development of the Tatar and Kazakh transducers to a related language.  While the Kazakh and Tatar transducers took years to reach their current production-quality (around 90\%) coverage of corpora, the Kumyk transducer only took a couple weeks to reach this level of quality.  This paper explores how the development of the Kumyk transducer benefited from knowledge gained from the development of the Tatar and Kazakh transducers.

@@

The Kazakh morphotactics were originally developed based on the Kyrgyz transducer, which was coauthored by a linguist who is fluent in and has a good linguistic knowledge of Kyrgyz together with a native speaker of Kyrgyz, and in consultation with another native speaker of Kyrgyz.  The initial developer of the Kazakh morphotactic was the same linguist, who is fluent in and has a good linguistic understanding Kazakh.  The morphotactics of Tatar were developed for the most part by a native Tatar speaker, who also worked to polish off the morphotactics of the Kazakh transducer.

@@

Each transducer's lexc source consists of lists of stems, with each stem pointing at a complex continuation lexicon containing the appropriate morphology for the type of stem.  As the compiled transducer parses the stems and morphology, tags are returned.

%FIXME: update for the three languages
The tagset for each transducer is designed to be compatible with the others.  Each transducer consists of about 120 separate tags, of which close to 20 cover the main parts of speech (noun, verb, adjective, adverb, postposition, interjection, etc.).  The remaining tags cover morphological subcategorisation for e.g.\ case, number, person, possession, transitivity, tense-aspect-mood, etc.  The tags are represented as multicharacter symbols, between less-than `<' and greater-than `>' symbols.  The tagset is quite extensive and still not entirely stabilised, so a full listing is not included here.  However, the tags are listed in the source code of the transducers, along with comments describing their usage.

@@

%FIXME: sources
We tested the coverage of the Kazakh and Tatar analysers over three separate domains: encyclopaedic text,\footnote{The following wikipedia dumps were used: kkwiki-20131006-pages-articles.xml.bz2, FIXME.} news,\footnote{All content from \url{http://www.azattyq.org/} for 2010 was used for Kazakh, as well as all content from 2005 to 2011 on \url{http://tat.tatar-inform.ru} for Tatar.} and religion.\footnote{We used a Kazakh bible translation available from \url{https://kkitap.net/} and a Tatar translation of the New Testament available from \url{http://ibt.org.ru}}  As there is currently no wikipedia in Kumyk, we tested only news and religion.\footnote{The bible corpus is from \url{http://ibt.org.ru/} and the news corpus consists of all Kumyk content from \url{http://sh-tavisi.etnosmi.ru/}.}

% Tatar, religion: tat.nt.ibt.txt, tat.quran.nughmani.txt.bz2
% Tatar, wp: tat.wikipedia.2013-02-25.txt.bz2
% Tatar, news: tat.news.2005-2011_300K-sentences.txt.bz2
% Kazakh, religion: kaz.bible.kkitap.txt.bz2, kaz.quran.altay.txt.bz2
% Kazakh, wp: kaz.wikipedia2.2013.txt.bz2
% Kazakh, news: kaz.rferl.2010.xml.bz2, kaz.rferl2.2012.txt
% Kumyk, religion: kum.nt.ibt.txt, (kum.genesis.ibt.txt)
% Kumyk, news: kum.yoldash.2013.txt, kum.articles.kumukia.txt.bz2

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Words} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia &  -  &  - \\
	& News & 3.2M & - \\
	& Religion & 577K & - \\\cline{2-4}
	& Average & - & 90.5\% \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 128K &  - \\
	%FIXME: the News corpus isn't RFERL, pretty sure; maybe we should test just on RFERL content?
	& News & 4.6M & - \\
	& Religion & 137K & - \\\cline{2-4}
	& Average & - & 89.0\% \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 287K &  91.4 $\pm$ 0.7 \\
	& Religion & 227K & 92.0 $\pm$ 1.1 \\\cline{2-4}
	& Average & - & 91.7 $\pm$ 0.7 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\end{table}

@@

%FIXME: footnote with urls
This list of surface forms was then analysed with the most recent version of the analyser, and each analysis was checked.  Where an analysis was erroneous, it was removed; where an analysis was missing, it was added.  This process gave us a `gold standard' morphologically analysed word list of 500 surface forms with their analyses.  The list is publically available for each language.%\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/apertium-kir/eval/ref.1000.txt}}

@@

One direction for future work is to develop transducers for more languages.  We have already constructed usable prototype transducers for three other Kypchak languages: Bashkir (N), Nogay (S), and Karakalpak (S).  Since our ability to develop production-ready transducers is limited by availability of resources, including corpora in the languages and native speaker consultants, the Western Kypchak languages (aside from Kumyk) have been more neglected by our team.  However, these language communities would benefit from computational tools for the language, and work on them may be bootstrapped from the existing transducers, so working on morphological transducers for these languages is also a priority.

>>>>

% FIXME: what are transducers used for ?

@@

\begin{map*}[htbp]
	\includegraphics[width=\textwidth]{map/map}
	\caption{The three sub-branches of Kypchak (North, South, West), roughly divided with black lines, showing the geographic distribution of the three languages for which transducers were developed (\texttt{tat}, \texttt{kaz}, \texttt{kum}).  Language codes are from ISO 639-3.}
	\label{map1}
\end{map*}

@@

The Kumyk transducer was developed starting at the beginning of October, 2013 as experiment to see how difficult it would be to extend lessons learned from the development of the Tatar and Kazakh transducers to a related language.  While the Kazakh and Tatar transducers took years of part-time work to reach their current production-quality (around 90\%) coverage of corpora, the Kumyk transducer only took a couple weeks to reach this level of quality.  This paper explores how the development of the Kumyk transducer benefited from knowledge gained from the development of the Tatar and Kazakh transducers.

@@

The Kazakh morphotactics were originally developed based on the Kyrgyz transducer, which was co-authored by a linguist who is fluent in and has a good linguistic knowledge of Kyrgyz together with a native speaker of Kyrgyz, and in consultation with another native speaker of Kyrgyz.  The initial developer of the Kazakh morphotactic was the same linguist, who is fluent in and has a good linguistic understanding Kazakh.  The morphotactics of Tatar were developed for the most part by a native Tatar speaker, who also worked to polish off the morphotactics of the Kazakh transducer.

@@

Each transducer's lexc source consists of lists of stems, with each stem pointing at a complex continuation lexicon containing the appropriate morphology for the type of stem.

%FIXME: update for the three languages
The tagset for each transducer is designed to be compatible with the others.  Each transducer consists of about 120 separate tags, of which close to 20 cover the main parts of speech (noun, verb, adjective, adverb, postposition, interjection, etc.).  The remaining tags cover morphological subcategorisation for e.g.\ case, number, person, possession, transitivity, tense-aspect-mood, etc.  The tags are represented as multicharacter symbols, between less-than \texttt{<} and greater-than \texttt{>} symbols.  The tagset is quite extensive and still not entirely stabilised, so a full listing is not included here.  However, the tags are listed in the source code of the transducers, along with comments describing their usage.

@@

%FIXME: sources
We tested the coverage of the Kazakh and Tatar analysers over three separate domains: encyclopaedic text,\footnote{The following Wikipedia dumps were used: kkwiki-20131006-pages-articles.xml.bz2, FIXME.} news,\footnote{All content from \url{http://www.azattyq.org/} for 2010 was used for Kazakh, as well as all content from 2005 to 2011 on \url{http://tat.tatar-inform.ru} for Tatar.} and religion.\footnote{We used a Kazakh bible translation available from \url{https://kkitap.net/} and a Tatar translation of the New Testament available from \url{http://ibt.org.ru}}  As there is currently no Wikipedia in Kumyk, we tested only news and religion.\footnote{The bible corpus is from \url{http://ibt.org.ru/} and the news corpus consists of all Kumyk content from \url{http://sh-tavisi.etnosmi.ru/}.}

% Kazakh, religion: kaz.bible.kkitap.txt.bz2, kaz.quran.altay.txt.bz2
% Kazakh, wp: kaz.wikipedia2.2013.txt.bz2
% Kazakh, news: kaz.rferl.2010.xml.bz2, kaz.rferl2.2012.txt

% Tatar, wp: tat.wikipedia.2013-02-25.txt.bz2
% Tatar, news: tat.news.2005-2011_300K-sentences.txt.bz2
% Tatar, religion: tat.nt.ibt.txt, tat.quran.nughmani.txt.bz2

% Kumyk, news: kum.yoldash.2013.txt, kum.articles.kumukia.txt.bz2
% Kumyk, religion: kum.nt.ibt.txt, (kum.genesis.ibt.txt)

@@

\begin{table}
\begin{center}
\begin{tabular}{llrr}
\toprule
\textbf{Language} & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) \\
\midrule
\multirow{4}{*}{Kazakh} & Wikipedia &  -  &  - \\
	& News & 3.2M & - \\
	& Religion & 848K & 92.7 $\pm$ 1.7 \\\cline{2-4}
	& Average & - &  \\
\midrule
\multirow{4}{*}{Tatar} & Wikipedia & 128K &  - \\
	& News & 4.6M & - \\
	& Religion & 205K & 88.7 $\pm$ 0.5 \\\cline{2-4}
	& Average & - &  \\
\midrule
% news = kum.yoldash.2013.txt
% religion = kum.genesis.ibt.txt, kum.nt.ibt.txt
\multirow{4}{*}{Kumyk} & Wikipedia & -- & -- \\
        & News & 287K &  91.4 $\pm$ 0.7 \\
	& Religion & 227K & 92.0 $\pm$ 1.1 \\\cline{2-4}
	& Average & - & 91.7 $\pm$ 0.7 \\
\bottomrule
\end{tabular}
 \caption{Corpora used for naïve coverage tests}
 \label{table:corpora}
\end{center}
\end{table}

@@

%FIXME: footnote with urls
This list of surface forms was then analysed with the most recent version of the analyser, and each analysis was checked.  Where an analysis was erroneous, it was removed; where an analysis was missing, it was added.  This process gave us a `gold standard' morphologically analysed word list of 1,000 surface forms with their analyses.  The list is publically available for each language.%\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/branches/apertium-kir/eval/ref.1000.txt}}

@@

One direction for future work is to develop transducers for more languages.  We have already constructed usable prototype transducers for three other Kypchak languages: Bashkir (N), Nogay (S), and Karakalpak (S).  Since our ability to develop production-ready transducers is limited by availability of resources, including corpora in the languages and native-speaker consultants, the Western Kypchak languages (aside from Kumyk) have been more neglected by our team.  However, these language communities would benefit from computational tools for their languages, and work on them may be bootstrapped from the existing transducers, so working on morphological transducers for these languages is also a priority.
